---
title: "Homework 2 - Problem 1: Random Walk with Momentum"
format:
  html:
    code-fold: false
    embed-resources: true
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## (a) Simulation function and input validation

The four rows of `directions` represent right, left, up, and down. Both
implementations below return a named vector containing the final coordinates.

```{r}
directions <- rbind(
  right = c(x = 1L, y = 0L),
  left  = c(x = -1L, y = 0L),
  up    = c(x = 0L, y = 1L),
  down  = c(x = 0L, y = -1L)
)

check_inputs <- function(n, p) {
  if (!is.numeric(n) || length(n) != 1L || !is.finite(n) ||
      n < 1 || n != floor(n)) {
    stop("n must be one integer greater than or equal to 1.")
  }
  if (!is.numeric(p) || length(p) != 1L || !is.finite(p) ||
      p < 0 || p > 1) {
    stop("p must be one number between 0 and 1.")
  }
  invisible(TRUE)
}
```

## (b) Two implementations

### Version 1: `for` loop

The first direction is uniformly sampled from the four possibilities. At each
later step, a Bernoulli trial determines whether to preserve the previous
direction; otherwise a new direction is sampled uniformly. The final position
is the column sum of all step vectors.

```{r}
momentum_walk1 <- function(n, p) {
  check_inputs(n, p)
  n <- as.integer(n)

  direction_ids <- integer(n)
  direction_ids[1] <- sample.int(4L, 1L)

  if (n >= 2L) {
    for (step in 2:n) {
      if (runif(1L) < p) {
        direction_ids[step] <- direction_ids[step - 1L]
      } else {
        direction_ids[step] <- sample.int(4L, 1L)
      }
    }
  }

  final_position <- colSums(directions[direction_ids, , drop = FALSE])
  setNames(as.integer(final_position), c("x", "y"))
}
```

### Version 2: `sapply`

This version uses `sapply()` to generate steps 2 through *n*. The variable
`previous_direction` stores the state that the next call needs. The first
direction is then joined to the generated directions and summed.

```{r}
momentum_walk2 <- function(n, p) {
  check_inputs(n, p)
  n <- as.integer(n)

  first_direction <- sample.int(4L, 1L)
  previous_direction <- first_direction

  remaining_directions <- if (n == 1L) {
    integer(0)
  } else {
    sapply(seq_len(n - 1L), function(step) {
      current_direction <- if (runif(1L) < p) {
        previous_direction
      } else {
        sample.int(4L, 1L)
      }
      previous_direction <<- current_direction
      current_direction
    })
  }

  direction_ids <- c(first_direction, remaining_directions)
  final_position <- colSums(directions[direction_ids, , drop = FALSE])
  setNames(as.integer(final_position), c("x", "y"))
}
```

The requested demonstrations are:

```{r}
momentum_walk1(10, 0.6)
momentum_walk2(10, 0.6)
momentum_walk1(1000, 0.6)
momentum_walk2(1000, 0.6)
```

## (c) Reproducibility

`set.seed()` resets R's pseudo-random-number generator to a known state. If the
same seed is set immediately before the same simulation, the function receives
the same sequence of random values and therefore returns the same result.

```{r}
set.seed(506)
walk_10_first <- momentum_walk1(10, 0.6)
set.seed(506)
walk_10_second <- momentum_walk1(10, 0.6)

set.seed(506)
walk_1000_first <- momentum_walk1(1000, 0.6)
set.seed(506)
walk_1000_second <- momentum_walk1(1000, 0.6)

data.frame(
  n = c(10, 1000),
  first_x = c(walk_10_first["x"], walk_1000_first["x"]),
  first_y = c(walk_10_first["y"], walk_1000_first["y"]),
  repeated_x = c(walk_10_second["x"], walk_1000_second["x"]),
  repeated_y = c(walk_10_second["y"], walk_1000_second["y"]),
  identical = c(
    identical(walk_10_first, walk_10_second),
    identical(walk_1000_first, walk_1000_second)
  ),
  row.names = NULL
)
```

## (d) Does momentum matter in three walks?

```{r}
distance_from_origin <- function(position) {
  sqrt(sum(position^2))
}

set.seed(506)
momentum_values_d <- c(0, 0.5, 0.9)
single_walk_results <- data.frame(
  p = momentum_values_d,
  t(sapply(momentum_values_d, function(p) momentum_walk1(1000, p)))
)
single_walk_results$distance <- apply(
  single_walk_results[c("x", "y")], 1L, distance_from_origin
)
single_walk_results
```

In these three simulations, the final distance generally grows with momentum.
However, each row is only one random walk, so a different seed can produce a
different ordering. A Monte Carlo study is more appropriate for describing
typical behavior.

## (e) Monte Carlo simulation

For each value of `p`, `replicate()` runs 1,000 independent walks. The code
calculates each final distance and then takes its mean.

```{r}
set.seed(506)
momentum_values <- c(0, 0.25, 0.5, 0.75, 0.9)

mean_distances <- sapply(momentum_values, function(p) {
  distances <- replicate(
    1000,
    distance_from_origin(momentum_walk1(1000, p))
  )
  mean(distances)
})

monte_carlo_results <- data.frame(
  p = momentum_values,
  mean_distance = round(mean_distances, 3)
)
monte_carlo_results
```

The mean final distance increases as `p` increases. Momentum creates positive
dependence between consecutive steps: long runs in one direction become more
common, so cancellation among steps is reduced. The increase is especially
large near `p = 1`.

## (f) Computational performance

The benchmarks use more repetitions for the shorter walk and fewer for the
longer walk so that the document remains practical to render.

```{r}
library(microbenchmark)

set.seed(506)
benchmark_1000 <- microbenchmark(
  loop = momentum_walk1(1000, 0.6),
  sapply = momentum_walk2(1000, 0.6),
  times = 100L
)

set.seed(506)
benchmark_100000 <- microbenchmark(
  loop = momentum_walk1(100000, 0.6),
  sapply = momentum_walk2(100000, 0.6),
  times = 10L
)

summary(benchmark_1000, unit = "ms")[c("expr", "median", "mean")]
summary(benchmark_100000, unit = "ms")[c("expr", "median", "mean")]
```

The timings printed above determine which implementation is faster on this
computer. Both versions still perform one state-dependent update per step, so
`sapply()` cannot fully vectorize this simulation. In addition, it repeatedly
calls an anonymous function and updates an enclosing variable. That overhead
usually makes the explicit loop faster, and the absolute timing gap becomes
more noticeable for 100,000 steps even if the speed ratio is similar.
