---
title: "Homework 2 - Problem 2: Campus Bus Arrivals"
format:
  html:
    code-fold: false
    embed-resources: true
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

## Monte Carlo simulation

The simulation below generates all 10,000 days at once. For each multi-hour
period, `rpois()` produces one value per hour and day, and `rowSums()` adds the
hourly values within each day. The Normal distributions use standard
deviations of $\sqrt{16}=4$ and $\sqrt{25}=5$. No loops are used.

```{r}
set.seed(506)

n_days <- 10000L

# Each row represents one simulated day.
passengers_6_to_8 <- rowSums(matrix(
  rpois(n_days * 3L, lambda = 4),
  nrow = n_days,
  ncol = 3L
))

passengers_9 <- rnorm(
  n_days,
  mean = 45,
  sd = sqrt(16)
)

passengers_10_to_3 <- rowSums(matrix(
  rpois(n_days * 6L, lambda = 15),
  nrow = n_days,
  ncol = 6L
))

passengers_4 <- rnorm(
  n_days,
  mean = 55,
  sd = sqrt(25)
)

passengers_5_to_10 <- rowSums(matrix(
  rpois(n_days * 6L, lambda = 9),
  nrow = n_days,
  ncol = 6L
))

daily_total <- (
  passengers_6_to_8 +
  passengers_9 +
  passengers_10_to_3 +
  passengers_4 +
  passengers_5_to_10
)

estimated_mean <- mean(daily_total)
monte_carlo_se <- sd(daily_total) / sqrt(n_days)

simulation_result <- data.frame(
  simulated_days = n_days,
  estimated_mean = estimated_mean,
  monte_carlo_se = monte_carlo_se,
  lower_95 = estimated_mean - 1.96 * monte_carlo_se,
  upper_95 = estimated_mean + 1.96 * monte_carlo_se
)

round(simulation_result, 3)
```

The Monte Carlo estimate is approximately **`r round(estimated_mean, 2)`
passengers per day**.

## Reasonableness check

Because expectations add, the theoretical daily mean is

$$
3(4) + 45 + 6(15) + 55 + 6(9) = 256.
$$

```{r}
theoretical_mean <- 3 * 4 + 45 + 6 * 15 + 55 + 6 * 9

data.frame(
  monte_carlo_estimate = round(estimated_mean, 3),
  theoretical_mean = theoretical_mean,
  simulation_error = round(estimated_mean - theoretical_mean, 3)
)
```

The simulation estimate is close to 256, as expected. Any small difference is
ordinary Monte Carlo sampling error and would tend to shrink if more days were
simulated.
